---
title: "Assignment6"
author: "Daphne Jacobsberg"
date: "10/26/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(censusapi)
library(mapview)
library(devtools)
library(leaflet)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r}
pums_2019_1yr <- getCensus(
  name = "acs/acs1/pums",
  vintage = 2019,
  region = "public use microdata area:*", 
  regionin = "state:06",
  vars = c(
    "SERIALNO",
    "SPORDER",
    "PWGTP",
    "WGTP",
    "NP",
    "HINCP",
    "RWAT",
    "YBL",
    "BLD",
    "TEN",
    "MV",
    "AGEP"
  )
)
```


```{r}
ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

sf_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME == "San Francisco")

sf_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[sf_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()
```

Here is a map of the PUMAs in the San Francisco County:
```{r}
leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data = sf_pumas,
    color = "gray",
    opacity = 0.7,
    fillOpacity = 0.3,
    weight = 1.5
    )
```

```{r}
sf_pums <-
  pums_2019_1yr %>% 
  mutate(
    PUMA = str_pad(public_use_microdata_area,5,"left","0"),
    YBL = as.numeric(YBL),
    HINCP = as.numeric(HINCP),
    AGEP = as.numeric(AGEP)
  ) %>% 
  group_by(SERIALNO) %>%
  arrange(AGEP) %>%
  summarize(
    AGEP = first(AGEP),
    YBL = first(YBL),
    HINCP = first(HINCP),
    TEN = first(TEN),
    MV = first(MV),
    BLD = first(BLD),
    PUMA = first(PUMA),
    WGTP = first(WGTP)
  ) %>%
  filter(PUMA %in% sf_pumas$PUMACE10) %>%
  filter(YBL %in% 1:3) %>%
  mutate(
    leadrisk = ifelse(
      (HINCP < 90000) &
        (AGEP %in% 0:5),
      1,
      0
    )
  )

sf_pums_factored <-
  sf_pums %>%
  mutate(
    building = BLD %>% 
      factor(
        levels = sf_pums$BLD %>% 
          unique() %>%
          as.numeric() %>% 
          sort()
      ),
    tenure = TEN %>%
      factor(
        levels = sf_pums$TEN %>% 
          unique() %>%
          as.numeric() %>% 
          sort()
      ),
    movein = MV %>%
      factor(
        levels = sf_pums$MV %>% 
          unique() %>%
          as.numeric() %>% 
          sort()
      ),
    puma = PUMA %>%
      factor(
        levels = sf_pums$PUMA %>% 
          unique() %>%
          sort()
      )
  )
```


```{r}
logit_model <- glm(
  leadrisk ~ building + tenure + movein + puma,
  family = quasibinomial(),
  data = sf_pums_factored
)

summary(logit_model)
```

Here is a random prediction for the current model:

```{r}
random_predict <-
  predict(logit_model, data.frame(sample_n(sf_pums_factored,1)), type = "response")

random_predict
```

Keep in mind, this value will change after every run.

```{r}
model_predict <-
  predict(logit_model, type = "response")
```

```{r}
sf_pums_binded <-
  sf_pums_factored %>%
  cbind(model_predict)
```

bld = 2
mv = 3 or 4
ten = 3 or 4

```{r}
summary_2x2 <-
  sf_pums_binded %>% 
  mutate(
    leadrisk = ifelse(
      leadrisk == 1, 
      "Yes (Leadrisk)", 
      "No (Predict)"
    )
  ) %>% 
  pull(leadrisk) %>% 
  table(model_predict > 0.1)

summary_2x2
```

rarely owned


2200 are correct
27 + 4 how many the models says need help
33 + 4 are how many actually need help
but only 4/37 get the help

get more accurate model and then
have inspections 30 days after kit receival at houses that need help

