---
title: "Assignment4"
author: "Daphne Jacobsberg"
date: "10/18/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(censusapi)
library(mapview)
library(devtools)
library(leaflet)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r}
#just to choose variable - data dictionary
acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )

#variable code = C15002
```


```{r}
race_categories <- c(
  "White alone",
  "Black or African American alone",
  "American Indian and Alaska Native alone",
  "Asian alone",
  "Native Hawaiian and Other Pacific Islander alone",
  "Some Other Race alone",
  "Two or more Races"
)


bay_educ_race <-
  1:7 %>% 
  map_dfr(function(x){
    
    getCensus(
      name = "acs/acs5",
      vintage = "2019",
      region = "county:085",
      regionin = "state:06",
      vars = paste0("group(C15002",LETTERS[x],")")
    ) %>% 
      select(ends_with("E") & -c(state, NAME)) %>% 
      summarize_all(sum) %>% 
      pivot_longer(
        everything(),
        names_to = "name",
        values_to = "estimate"
      ) %>% 
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      separate(
        label,
        into = c(NA, NA, NA, "education"),
        sep = "!!"
      ) %>% 
      select(-name) %>% 
      filter(!is.na(education)) %>% 
      group_by(education) %>%
      summarize(across(everything(), sum)) %>%
      mutate(
        race = race_categories[x]
      )
  })



```

The graph below shows the educational attainment level of the Santa Clara 25 years or older population by race. Keeping in mind how developed and high-end Santa Clara county is, I did not find it surprising that the "Bachelor's degree or higher" was the most common. In fact, the number of individuals increases as you go down each category. Furthermore, looking at the distributions within, White alone and Asian alone were by far the most common and followed the trend described above. On the other hand, the Some Other Race alone category was at it highest in the lowest educational level and at it lowest in the highest educational level. This shows a significant drop out pattern which is very unfortunate.

```{r}
plot_stacked <-
  bay_educ_race %>% 
    group_by(education, race) %>% 
    summarize(estimate = sum(estimate)) %>% 
    ggplot() +
    geom_bar(
      aes(
        x = education %>% factor(levels = rev(c("Less than high school diploma", "High school graduate (includes equivalency)", "Some college or associate's degree", "Bachelor's degree or higher"))),
        y = estimate,
        fill = race %>% factor(levels = rev(unique(bay_educ_race$race)))
      ),
      stat = "identity",
      position = "stack"
    ) +
    labs(
      x = "Level of Educational Attainment",
      y = "Number of individuals",
      title = "Santa Clara educational attainment by race",
      fill = "Race of individual"
    ) +
    coord_flip() +
    theme(
      legend.position = "bottom",
      legend.direction = "vertical"
    )  +
    guides(
      fill = guide_legend(
        reverse = T
      )
    )

plot_stacked
```

While displaying the same data, the filled graph below is also interesting. It is especially useful when trying to understand the proportions of each race category within the educational levels. As it is a proportion, the x-axis scale becomes from 0 to 1 which makes it easier to interpret the data and draw conclusions from it. In almost all cases, White alone constitutes  50% of the population; and combined with Asian alone, 75% of the population. Looking specifically into the Black or African American alone category, you can see an increasing trend up to "Some college or associate's degree" but then a significant drop when it comes to completing a "Bachelor's degree or higher". Investigating the contributing factors could also be an interesting future exploration. Overall, we can see there is a clear disproportionate education attainment by race in the county.

```{r}
bay_race_total <-
  bay_educ_race %>% 
  group_by(race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(education = "Total")

plot_filled <-
  bay_educ_race %>% 
    group_by(education, race) %>% 
    summarize(estimate = sum(estimate)) %>% 
    rbind(bay_race_total) %>% 
    ggplot() +
    geom_bar(
      aes(
        x = education %>% factor(levels = rev(c("Total","Less than high school diploma", "High school graduate (includes equivalency)", "Some college or associate's degree", "Bachelor's degree or higher"))),
        y = estimate,
        fill = race %>% factor(levels = rev(unique(bay_educ_race$race)))
      ),
      stat = "identity",
      position = "fill"
    ) +
    labs(
      x = "Level of Educational Attainment",
      y = "Proportion of individuals",
      title = "Santa Clara educational attainment by race",
      fill = "Race of individual"
    ) +
    coord_flip() +
    theme(
      legend.position = "bottom",
      legend.direction = "vertical"
    )  +
    guides(
      fill = guide_legend(
        reverse = T
      )
    )

plot_filled
```



6.5 assignment

```{r}
acs_vars_2019 <-
  listCensusMetadata(
    name = "2019/acs/acs1",
    type = "variables"
  )


latin_categories <- c(
  "Latinx",
  "White, not latinx",
  "Not white, not latinx"
)
```

```{r}
bay_educ_white <-
    getCensus(
      name = "acs/acs1",
      vintage = "2019",
      region = "county:085",
      regionin = "state:06",
      vars = "group(C15002H)",
    ) %>% 
      select(ends_with("E") & -c(state, NAME)) %>% 
      summarize_all(sum) %>% 
      pivot_longer(
        everything(),
        names_to = "name",
        values_to = "estimate"
      ) %>% 
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      separate(
        label,
        into = c(NA, NA, NA, "education"),
        sep = "!!"
      ) %>% 
      select(-name) %>% 
      filter(!is.na(education)) %>% 
      group_by(education) %>%
      summarize(across(everything(), sum)) %>%
      mutate(
        race = "White, not latinx"
      )


bay_educ_latinx <-
    getCensus(
      name = "acs/acs1",
      vintage = "2019",
      region = "county:085",
      regionin = "state:06",
      vars = "group(C15002I)",
    ) %>% 
      select(ends_with("E") & -c(state, NAME)) %>% 
      summarize_all(sum) %>% 
      pivot_longer(
        everything(),
        names_to = "name",
        values_to = "estimate"
      ) %>% 
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      separate(
        label,
        into = c(NA, NA, NA, "education"),
        sep = "!!"
      ) %>% 
      select(-name) %>% 
      filter(!is.na(education)) %>% 
      group_by(education) %>%
      summarize(across(everything(), sum)) %>%
      mutate(
        race = "Latinx"
      )

bay_educ_race_1yr <-
    getCensus(
      name = "acs/acs1",
      vintage = 2019,
      region = "county:085",
      regionin = "state:06",
      vars = "group(C15002)"
      ) %>% 
      select(!c(GEO_ID,state, NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019 %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA, "education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>% 
      mutate(
        ethnicity = "Total",
        education = case_when(
          education %in% c("Less than 9th grade", "9th to 12th grade, no diploma") ~ "Less than high school diploma",
          education == "High school graduate (includes equivalency)" ~ education,
          education %in% c("Some college, no degree", "Associate's degree") ~ "Some college or associate's degree",
          education %in% c("Bachelor's degree", "Graduate or professional degree") ~ "Bachelor's degree or higher"
          )
      )%>%
      group_by(education, county, ethnicity) %>%
      summarize_all(sum)


bay_educ_total <-
  bay_educ_race_1yr %>% 
  group_by(education) %>% 
  summarize(estimate = sum(estimate)) %>%
  mutate(
    race = "Total"
  )

bay_educ_neither <-
  bay_educ_total %>%
  mutate(
    neither = estimate - bay_educ_white$estimate - bay_educ_latinx$estimate,
    race = "Not white, not latinx"
  )%>%
  select(-estimate)%>%
  rename(estimate = neither)
  

```

The two graphs below show the Santa Clara county educational attainment level by ethnicity. It is very clear that the Latinx population is less likely to attain the same level of education as the non latinx population. Despite being less than a third of the total population, Latinxs are more than half of the population that have completed "less than high school diploma". Looking at the other two categories, it is interesting to point out that proportionally, they are pretty much equal at all levels except for "Less than high school diploma" in which the "White, not latinx" population is significantly smaller.


```{r}
plot_stacked_latinx <-
  bay_educ_neither %>% 
    #group_by(education, race) %>% 
    #summarize(estimate = sum(estimate)) %>% 
    rbind(bay_educ_white) %>%
    rbind(bay_educ_latinx) %>%
    ggplot() +
    geom_bar(
      aes(
        x = education %>% factor(levels = rev(c("Less than high school diploma", "High school graduate (includes equivalency)", "Some college or associate's degree", "Bachelor's degree or higher"))),
        y = estimate,
        fill = race
      ),
      stat = "identity",
      position = "stack"
    ) +
    labs(
      x = "Level of Educational Attainment",
      y = "Number of individuals",
      title = "Santa Clara educational attainment by ethnicity",
      fill = "Ethnicity of individual"
    ) +
    coord_flip() +
    theme(
      legend.position = "bottom",
      legend.direction = "vertical"
    )  +
    guides(
      fill = guide_legend(
        reverse = T
      )
    )

plot_stacked_latinx
```


```{r}
plot_latinx <-
  bay_educ_neither %>% 
    group_by(education, race) %>% 
    summarize(estimate = sum(estimate)) %>% 
    rbind(bay_educ_white) %>%
    rbind(bay_educ_latinx)


bay_ethnic_total <-
  plot_latinx %>%
  group_by(race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(education = "Total")
  
  
plot_filled_latinx <-
  plot_latinx %>% 
    rbind(bay_ethnic_total) %>%
    ggplot() +
    geom_bar(
      aes(
        x = education %>% factor(levels = rev(c("Total", "Less than high school diploma", "High school graduate (includes equivalency)", "Some college or associate's degree", "Bachelor's degree or higher"))),
        y = estimate,
        fill = race
      ),
      stat = "identity",
      position = "fill"
    ) +
    labs(
      x = "Level of Educational Attainment",
      y = "Number of individuals",
      title = "Santa Clara educational attainment by ethnicity",
      fill = "Ethnicity of individual"
    ) +
    coord_flip() +
    theme(
      legend.position = "bottom",
      legend.direction = "vertical"
    )  +
    guides(
      fill = guide_legend(
        reverse = T
      )
    )

plot_filled_latinx
```




```{r}
#7pts attempt

bay_educ_white_loop <-
  2006:2019 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs1",
      vintage = x,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = "group(C15002H)"
      ) %>% 
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019 %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA, "education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>% 
      mutate( race = "White, not latinx") %>%
      group_by(education, county, race) %>%
      summarize_all(sum)
  })


bay_educ_latinx_loop <-
  2006:2019 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs1",
      vintage = x,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = "group(C15002I)"
      ) %>% 
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019 %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA, "education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>% 
      mutate( race = "White, not latinx") %>%
      group_by(education, race, county) %>%
      summarize_all(sum)
  })


bay_educ_total_loop <-
  2006:2019 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs1",
      vintage = x,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = "group(C15002)"
      ) %>% 
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019 %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA, "education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>% 
      mutate(
        ethnicity = "Total",
        education = case_when(
          education %in% c("Less than 9th grade", "9th to 12th grade, no diploma") ~ "Less than high school diploma",
          education == "High school graduate (includes equivalency)" ~ education,
          education %in% c("Some college, no degree", "Associate's degree") ~ "Some college or associate's degree",
          education %in% c("Bachelor's degree", "Graduate or professional degree") ~ "Bachelor's degree or higher"
          )
      )%>%
      group_by(education, county, ethnicity) %>%
      summarize_all(sum)
  })
```


```{r}
